---
title: "ST 412/512 Lab 8: Two-way ANOVA"
format: html
message: false
warning: false
---

## Objectives

- Create an interaction plot.
- Estimate contrasts.

## 1. Setup

Download the `Lab8Anova.r` script from Canvas and open it in RStudio. Load the required packages. The `dplyr` package offers many useful data-manipulation functions.

```{r}
library(Sleuth3)
library(ggplot2)
library(dplyr)
```

## 2. Create the Data Frame

Create the `seaweed` data frame and add a logit-transformed response:

```{r}
seaweed <- case1301
seaweed$logitCover <- with(case1301, log(Cover/(100 - Cover)))
```

## 3. Interaction Plot

An interaction plot in a two-way ANOVA displays group means on the vertical axis and levels of one factor on the horizontal axis. The other factor is represented by different line types or plotting symbols.

### (a) Using `interaction.plot()`

```{r}
with(seaweed, interaction.plot(Block, Treat, response = logitCover))
```

> The first argument specifies the factor for the horizontal axis. Typically, choose the factor with the most levels (here, `Block`).

### (b) Using `dplyr` and `ggplot2`

First, calculate the mean response for each Block–Treatment combination.

1. **Group the Data:**

```{r}
seaweed1 <- group_by(seaweed, Block, Treat)
```

Printing `seaweed1` shows a tibble with grouping information (first by `Block`, then by `Treat`).

2. **Summarize the Data:**

```{r}
seaweed2 <- summarize(seaweed1, Trt_mean = mean(logitCover))
seaweed2
```

The result is a tibble with grouping only by `Block`.

3. **Plot the Data:**

```{r}
ggplot(data = seaweed2, aes(x = Block, y = Trt_mean, shape = Treat)) +
  geom_point() +
  geom_line(aes(group = Treat))
```

### (d) Optional: Using the Pipe Operator

The above two-step process can be streamlined with the pipe operator `%>%`:

```{r}
seaweed %>%
  group_by(Block, Treat) %>%
  summarize(Trt_mean = mean(logitCover)) -> seaweed2
```

## 4. Estimating Contrasts

We now estimate contrasts to assess seaweed grazer effects. For example, the **Large Fish contrast** is given by:

$$
\frac{1}{2}\mu_{LfF} - \frac{1}{2}\mu_{Lf} + \frac{1}{2}\mu_{fF} - \frac{1}{2}\mu_{f}
$$

Contrasts are linear combinations of means with coefficients that sum to 0. Here, the coefficients for the Large Fish contrast are:

$$
\frac{1}{2},\ -\frac{1}{2},\ \frac{1}{2},\ -\frac{1}{2},\ 0,\ 0
$$

corresponding to treatment means $\mu_{LfF}$, $\mu_{Lf}$, $\mu_{fF}$, $\mu_{f}$, $\mu_{L}$, and $\mu_{C}$.

### (a) Estimating Contrasts via Sample Means

1. **Calculate Sample Means and Sizes:**

```{r}
summary(seaweed$Treat)
Means <- with(seaweed, unlist(lapply(split(logitCover, Treat), mean)))
names(Means) <- names(summary(seaweed$Treat))
Means
```

2. **Estimate $\sigma$:**

Fit the model to obtain $\sigma$:

```{r}
model <- lm(logitCover ~ Treat + Block, data = seaweed)
summary(model)
s <- summary(model)$sigma
s
```

### (b) Estimate the Large Fish Contrast and Its Standard Error

```{r}
ghat <- 0.5 * Means["LfF"] - 0.5 * Means["Lf"] + 0.5 * Means["fF"] - 0.5 * Means["f"]
SE_ghat <- s * sqrt((0.5)^2 / 16 + (-0.5)^2 / 16 + (0.5)^2 / 16 + (-0.5)^2 / 16)
```

Compare these calculations with the estimates in Display 13.13 from the lab manual.

### (c) Construct a 95% Confidence Interval

The degrees of freedom for the $t$-quantile are from the residual degrees of freedom in the model.

```{r}
ci_lower <- ghat - qt(0.975, 83) * SE_ghat
ci_upper <- ghat + qt(0.975, 83) * SE_ghat
ci_lower
ci_upper
```

### (d) Optional: Estimating All Five Contrasts Simultaneously

Redefine the contrasts attribute of the factor variable `Treat` using a contrast matrix $ M $.

1. **Create the Matrix $ M $:**

```{r}
M <- matrix(c(0, 0, -1/2, 1/2, -1/2, 1/2,
              -1/2, -1/2, 1/2, 0, 1/2, 0,
              1/3, -1/3, -1/3, -1/3, 1/3, 1/3,
              1, -1, 1/2, 1/2, -1/2, -1/2,
              0, 0, -1, 1, 1, -1), nrow = 6, ncol = 5)
M
```

2. **Set the Contrasts for `Treat`:**

```{r}
contrasts(seaweed$Treat) <- M %*% solve(t(M) %*% M)
```

This instructs R to reparameterize the six groups in terms of five linearly independent contrasts.

3. **Estimate Parameters and Confidence Intervals:**

```{r}
model_contrasts <- lm(logitCover ~ Treat + Block, data = seaweed)
summary(model_contrasts)$coefficients
confint(model_contrasts)
```

Verify that these estimates agree with Display 13.13.

4. **Restore the Original Contrasts:**

```{r}
Indicators <- contr.treatment(6)
colnames(Indicators) <- names(Means)[-1]
contrasts(seaweed$Treat) <- Indicators
```

The second command assigns names to the indicator variables for clarity.

