---
title: "Generalized Linear Regression Homework 2"
author: "Brian Cervantes Alvarez"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format: OSULatexTheme-pdf
editor:
  render-on-save: true
message: false
warning: false
---

# Problem 1

**Consider the Donner Party example.** Answer the following questions based on the R output on Page 10 of Lecture 5 and the following covariance matrix of the estimated coefficient vector

$$
\begin{pmatrix}
\text{Covariances} & \text{Intercept} & \text{SexFemale} & \text{Age} \\
\text{Intercept} & 1.92338373 & 0.63080437 & -0.04670471 \\
\text{SexFemale} & 0.63080437 & 0.57073339 & -0.00823197 \\
\text{Age} & -0.04670471 & -0.00823197 & 0.001390134 \\
\end{pmatrix}
$$

The **logistic regression model** is

$$
\text{logit}(p) = \beta_0 + \beta_1 \times \text{SexFemale} + \beta_2 \times \text{Age}
$$

with estimated coefficients

- $\beta_0 = 1.63312$
- $\beta_1 = 1.59729$
- $\beta_2 = -0.07820$

{{< pagebreak>}}

## Part A

For a 50-year-old female

- $\text{SexFemale} = 1$
- $\text{Age} = 50$

First, determine **Linear Predictor** $\eta$

$$
\begin{aligned}
\eta &= \beta_0 + \beta_1 \times \text{SexFemale} + \beta_2 \times \text{Age} \\
&= 1.63312 + 1.59729 \times 1 + (-0.07820) \times 50 \\
&= 1.63312 + 1.59729 - 3.910 \\
&= -0.67959
\end{aligned}
$$

Second, calculate the **Survival Probability** $p$

$$
\begin{aligned}
p = \frac{e^\eta}{1 + e^\eta} = \frac{e^{-0.67959}}{1 + e^{-0.67959}} \approx \frac{0.5068}{1 + 0.5068} \approx 0.3365
\end{aligned}
$$

Then we can find the **Variance & Standard Error** of $\eta$
$$
\begin{aligned}
\text{Var}(\eta) &= \text{Var}(\beta_0) + \text{Var}(\beta_1) \times (\text{SexFemale})^2 + \text{Var}(\beta_2) \times (\text{Age})^2 \\
&\quad + 2 \times \text{Cov}(\beta_0, \beta_1) \times \text{SexFemale} + 2 \times \text{Cov}(\beta_0, \beta_2) \times \text{Age} + 2 \times \text{Cov}(\beta_1, \beta_2) \times \text{SexFemale} \times \text{Age} \\
&= 1.92338373 + 0.57073339 \times 1^2 + 0.001390134 \times 50^2 \\
&= 1.73739286 \\
\text{SE}(\eta) &= \sqrt{\text{Var}(\eta)} = \sqrt{1.73739286} \approx 1.318
\end{aligned}
$$

Penultimate, the **95% Confidence Interval** for $\eta$. Using $z_{0.975} = 1.96$, we can find the lower and upper bounds as follows
$$
\begin{aligned}
\text{Lower limit} &= \eta - 1.96 \times \text{SE}(\eta) = -0.67959 - 1.96 \times 1.318 \approx -0.67959 - 2.583 \approx -3.263 \\
\text{Upper limit} &= \eta + 1.96 \times \text{SE}(\eta) = -0.67959 + 1.96 \times 1.318 \approx -0.67959 + 2.583 \approx 1.903 \\
\end{aligned}
$$

Ultmately, we can find the 95% Confidence Interval for $p$
$$
\begin{aligned}
p_{Lower} &= \frac{e^{-3.263}}{1 + e^{-3.263}} \approx \frac{0.0383}{1 + 0.0383} \approx 0.0369 \\
p_{Upper} &= \frac{e^{1.903}}{1 + e^{1.903}} \approx \frac{6.703}{1 + 6.703} \approx 0.870 \\
95\% \text{ Confidence Interval} &= (0.0369, 0.870)
\end{aligned}
$$

{{< pagebreak>}}

## Part B

Let's first find the **Log-Odds** $\Delta\eta$. I did the following,

$$
\begin{aligned}
\Delta\eta &= \left( \beta_0 + \beta_1 \times 0 + \beta_2 \times 40 \right) - \left( \beta_0 + \beta_1 \times 1 + \beta_2 \times 35 \right) \\
&= \beta_2 \times (40 - 35) - \beta_1 \\
&= (-0.07820) \times 5 - 1.59729 \\
&= -0.391 - 1.59729 \\
&= -1.98829
\end{aligned}
$$

Next, we can determine the **Odds Ratio** by,

$$
\begin{aligned}
\text{OR} &= e^{\Delta\eta} = e^{-1.98829} \approx 0.1369
\end{aligned}
$$

Then, we can compute the **Variance** & **Standard Error** of $\Delta\eta$

$$
\begin{aligned}
\text{Var}(\Delta\eta) &= \text{Var}(-\beta_1 + 5\beta_2) \\
&= \text{Var}(\beta_1) + 25 \times \text{Var}(\beta_2) - 2 \times 5 \times \text{Cov}(\beta_1, \beta_2) \\
&= 0.57073339 + 25 \times 0.001390134 - 10 \times (-0.00823197) \\
&= 0.57073339 + 0.03475335 + 0.0823197 \\
&= 0.68780644 \\
\text{SE}(\Delta\eta) &= \sqrt{\text{Var}(\Delta\eta)} = \sqrt{0.68780644} \approx 0.829
\end{aligned}
$$

After that, we compute the 95% Confidence Interval for $\Delta\eta$
$$
\begin{aligned}
\text{Lower limit} &= \Delta\eta - 1.96 \times \text{SE}(\Delta\eta) = -1.98829 - 1.96 \times 0.829 \approx -1.98829 - 1.62484 \approx -3.61313 \\
\text{Upper limit} &= \Delta\eta + 1.96 \times \text{SE}(\Delta\eta) = -1.98829 + 1.96 \times 0.829 \approx -1.98829 + 1.62484 \approx -0.36345 \\
\end{aligned}
$$

Finally, we simply determine 95% Confidence Interval for the Odds Ratio as follows,
$$
\begin{aligned}
\text{OR}_{Lower} &= e^{-3.61313} \approx 0.027 \\
\text{OR}_{Upper} &= e^{-0.36345} \approx 0.695 \\
95\% \text{ Confidence Interval} &= (0.027, 0.695)
\end{aligned}
$$


{{< pagebreak>}}

# Problem 2

The data below are the launch temperatures *(degrees Fahrenheit)* and an indicator of O-ring failures for 24 space shuttle launches prior to the space shuttle Challenger disaster of January 27, 1986.

| Temperature | Failure | Temperature | Failure | Temperature | Failure |
|-------------|---------|-------------|---------|-------------|---------|
| 53          | Yes     | 68          | No      | 75          | No      |
| 56          | Yes     | 69          | No      | 75          | Yes     |
| 57          | Yes     | 70          | No      | 76          | No      |
| 63          | No      | 70          | Yes     | 76          | No      |
| 66          | No      | 70          | Yes     | 78          | No      |
| 67          | No      | 70          | Yes     | 79          | No      |
| 67          | No      | 72          | No      | 80          | No      |
| 67          | No      | 73          | No      | 81          | No      |


{{< pagebreak>}}


## Part A

We are given data on launch temperatures and O-ring failures. Let’s proceed with a logistic regression of `Failure` (binary: 1 for "Yes", 0 for "No") on `Temperature`.

1. Encode `Failure` as 1 for "Yes" and 0 for "No".
2. Fit the logistic regression model where the probability of failure is modeled as a function of `Temperature`.

   The estimated coefficients are:
   - Intercept ($\hat{\beta}_0$): 10.39366, SE: 5.52701
   - Temperature coefficient ($\hat{\beta}_1$): -0.16056, SE: 0.08014

Therefore,

- $\hat{\beta}_0 = 10.39366$, SE = 5.52701
- $\hat{\beta}_1 = -0.16056$, SE = 0.08014

{{< pagebreak>}}
```{r}
#| eval: true
shuttleDs <- data.frame(
  Temperature = c(53, 56, 57, 63, 66, 67, 67, 67, 68, 69, 
                  70, 70, 70, 70, 72, 73, 75, 75, 76, 76, 78, 79, 80, 81),
  Failure = c(1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 
              1, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0)
)
# Fit the logistic regression model
logisticModel <- glm(Failure ~ Temperature, family = binomial, 
                     data = shuttleDs)
summary(logisticModel)
```

{{< pagebreak>}}

## Part B

We will now test whether the coefficient of `Temperature` is 0 using Wald's test.

Let's define our hypotheses as follows,

   - Null Hypothesis ($H_0$): $\beta_1 = 0$
   - Alternative Hypothesis ($H_a$): $\beta_1 < 0$ (temperature decreases odds of failure)
   
Now, calculate Wald's test statistic
$$ Z = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)} = \frac{-0.16056}{0.08014} \approx -2.004 $$

```{r}
beta1 <- coef(summary(logisticModel))["Temperature", "Estimate"]
seBeta1 <- coef(summary(logisticModel))["Temperature", "Std. Error"]
Z <- beta1 / seBeta1
pVal <- pnorm(Z)
Z
pVal
```

Then, we can compute the one-sided p-value for $Z = -2.004$
$$ p \approx 0.02255718 $$

Thus, $p = 0.02255718 < 0.05$, we reject $H_0$. Hence, there is evidence that temperature negatively affects the odds of failure.

{{< pagebreak>}}

## Part C

We will now compute a 95% confidence interval for the temperature coefficient.
$$ \text{Lower limit} = \hat{\beta}_1 - 1.96 \times SE(\hat{\beta}_1) = -0.16056 - 1.96 \times 0.08014 \approx -0.3177 $$
$$ \text{Upper limit} = \hat{\beta}_1 + 1.96 \times SE(\hat{\beta}_1) = -0.16056 + 1.96 \times 0.08014 \approx -0.0034 $$

95% Confidence Interval for $\beta_1$: (-0.3177, -0.0034)

{{< pagebreak>}}

## Part D

Let’s interpret the effect of temperature on the odds of O-ring failure.

Thus, let us look at the odds ratio, where for a one-degree increase in temperature leads to,
$$ \text{OR} = e^{\hat{\beta}_1} = e^{-0.16056} \approx 0.8517 $$

Its 95% confidence interval is,

   - Lower limit: $e^{-0.3177} \approx 0.7276$
   - Upper limit: $e^{-0.0034} \approx 0.9966$

What does this mean? Well,

   - Each one-degree increase in temperature reduces the odds of O-ring failure by approximately 14.83%.
   - The true odds reduction is between 0.34% and 27.24%.

{{< pagebreak>}}

## Part E

Finally, we compute the estimated probability of failure at 31°F, where
$$ 
\begin{aligned}
\eta &= 10.39366 + (-0.16056) \times 31 \approx 5.4353 \\
p &= \frac{e^{5.4353}}{1 + e^{5.4353}} \approx 0.9956  
\end{aligned}
$$


Which means that the estimated probability of failure at 31°F is about 0.9956 (99.56%)