---
title: "ST557: Final Project"
author: 
    - "Brian Cervantes Alvarez"
date: "12-02-2023"
format: OSULatexTheme-pdf
execute: 
  warning: false
  message: false
editor: visual
---

# Read and Glance at Wine Datasets

```{r}
library(tidyverse)


# Read in the wine datasets
redWine <- read_csv("winequality-red.csv")
whiteWine <- read_csv("winequality-white.csv")

# Confirm the structure of the data given the description file
glimpse(redWine)
glimpse(whiteWine)

# Run a quick summary for both datasets
summary(redWine)
summary(whiteWine)
```
```{r}
# Load necessary libraries
library(tidyverse)


# Add a 'wine_type' column to identify the wine type
redWine$wineType <- "red"
whiteWine$wineType <- "white"

# Combine the datasets
wineDs <- bind_rows(redWine, whiteWine)

# Rename columns for better readability and consistency
wineDs <- wineDs %>% 
  dplyr::mutate(fixedAcidity = `fixed acidity`,
         volatileAcidity = `volatile acidity`,
         citricAcid = `citric acid`,
         residualSugar = `residual sugar`,
         freeSulfurDioxide = `free sulfur dioxide`,
         totalSulfurDioxide = `total sulfur dioxide`,
         wineType = factor(wineType, levels = c("red", "white"))) %>%
  dplyr::select(-c(`fixed acidity`,
            `volatile acidity`,
            `citric acid`,
            `residual sugar`,
            `free sulfur dioxide`,
            `total sulfur dioxide`))

# Display the column names of the combined dataset
names(wineDs)

# Display a concise summary of the structure of the dataset
glimpse(wineDs)

# Check for missing values in the dataset
sum(is.na(wineDs))

```

# Goal 1: Distinguish white wine from red wine

## Mean Vectors Comparison:

### Calculate the mean vectors for red and white wines separately for each of the 11 chemical attributes.

```{r}
# Calculate the mean vectors for red and white wines separately for each of the 11 chemical attributes
meanVectors <- wineDs %>% 
  group_by(wineType) %>%
  summarize_all(mean)

# Display the mean vectors
print("Mean Vectors:")
print(meanVectors)


meanVectors %>%
  ggplot(aes(x = wineType))
```



### Perform a statistical test ANOVA to determine if there is a significant difference in mean vectors between red and white wines.

The MANOVA results indicate that there is a significant difference in mean vectors between red and white wines for the specified chemical attributes. Here's a breakdown of the key information from the output:

Pillai's Trace: The value is 0.86158, which is a measure of the proportion of variance in the dependent variables explained by the independent variable (wine type). A value close to 1 indicates a strong effect.

Approximate F-statistic: The value is 3669.6, and the associated p-value is very close to zero (< 2.2e-16). This F-statistic is used to test the overall hypothesis that there are no differences between the mean vectors of the groups.

Degrees of Freedom: There is 1 degree of freedom for the wine type factor and 6495 degrees of freedom for residuals.

Pr(>F): The p-value associated with the F-statistic. In this case, it is highly significant (much less than 0.05), suggesting that there are significant differences in mean vectors between red and white wines.

The "***" next to the p-value indicates that the difference is highly significant.

In summary, based on the MANOVA results, there is strong evidence to reject the null hypothesis that there is no difference in mean vectors between red and white wines for the specified chemical attributes.

```{r}
wineManova <- manova(cbind(chlorides, density, pH, sulphates, 
                           alcohol, fixedAcidity, volatileAcidity, 
                           citricAcid, residualSugar, freeSulfurDioxide,
                           totalSulfurDioxide) ~ wineType, data = wineDs) 

# Print the summary of the MANOVA
summary(wineManova)
```



### The attributes with significant differences in mean vectors are the ones that differ the most between red and white wine.


### Perform a MANOVA To 




{{< pagebreak >}}

### b) Classification:

Split the dataset into training and testing sets.
Train a classification model (e.g., logistic regression, decision tree, or random forest) using the 11 chemical attributes as features and the wine type (red or white) as the target variable.
Evaluate the model's performance using metrics such as accuracy, precision, recall, and F1-score.
The classification rule would be based on the learned weights or feature importance from the model.

{{< pagebreak >}}

### c) Clustering:

Use hierarchical or k-means clustering to cluster the wines based on the 11 chemical attributes.
Choose a suitable distance measure (e.g., Euclidean distance) based on the nature of the data.
Validate the number of clusters using methods like the elbow method (for k-means) or dendrogram (for hierarchical clustering).
Evaluate how well the clusters reflect the red/white classification using metrics like silhouette score or cluster purity.

{{< pagebreak >}}


## Goal 2: Better understand which of these variables is most important to wine quality (Focus on red wine)

### a) Mean Vectors Comparison:**

   - Similar to the previous goal, calculate mean vectors for each of the 11 chemical attributes, but this time, group the wines based on quality scores (Low, Medium, High).
   - Perform a statistical test to determine if there is a significant difference in mean vectors between wines of different quality scores.
   - Identify which attributes show the most significant differences across quality categories.

{{< pagebreak >}}

### b) Classification/Prediction:**

   - Split the dataset into training and testing sets.
   - Use regression models (e.g., linear regression) to predict wine quality based on the 11 chemical attributes.
   - Evaluate the model's performance using metrics like Mean Squared Error (MSE) or R-squared.

{{< pagebreak >}}

### c) PCA Analysis:**

   - Perform Principal Component Analysis (PCA) on the 11 chemical attributes.
   - Plot the wines color-coded by quality score using the first two principal components.
   - Decide whether to standardize the data before PCA based on the scale of the original features. Standardization is generally recommended for PCA.
   - Create a new rule for predicting wine quality using only the scores for the first two principal components.
   - Compare the performance of this new classifier with the one using all predictor variables, using metrics like MSE or R-squared.

{{< pagebreak >}}









```{r}
#| echo: false
#| eval: false

# Visualize 
for (col in names(redWine)) {
  p <- redWine %>% ggplot(aes(x = !!sym(col))) +
    labs(title = paste0("Histogram Plot for: ", col)) +
    geom_histogram(,
                   bins = 30, 
                   fill = "#722F37",
                   color = "black", 
                   alpha = 0.7) +
    theme_minimal() 
  
  print(p)
}


# Visualize 
for (col in names(whiteWine)) {
  p <- redWine %>% ggplot(aes(x = !!sym(col))) +
    labs(title = paste0("Histogram Plot for: ", col)) +
    geom_histogram(,
                   bins = 30, 
                   fill = "#722F37",
                   color = "black", 
                   alpha = 0.7) +
    theme_minimal() 
  
  print(p)
}

```

