---
title: "Lab 4: ST 412/512"
author: "Brian Cervantes Alvarez"
format: 
  OSULatexTheme-pdf:
    toc: true
---

{{< pagebreak >}}

# Inference on Multiple Linear Regression in R

**Lab Goals:**

1. Perform F-tests for model comparison
2. Calculate and interpret confidence intervals for mean responses
3. Create prediction intervals for individual observations

{{< pagebreak >}}

## F-tests for Model Comparison

### Nested Model F-test
When comparing two nested models (where one is a simplified version of the other), we use an extra sum of squares F-test. In R:

```{r}
#| message: false
library(Sleuth3)

# Continuous intensity model
fit_continuous <- lm(Flowers ~ Intensity + Time, data = case0901)

# Categorical intensity model
fit_indicator <- lm(Flowers ~ factor(Intensity) + Time, data = case0901)

# Model comparison
anova(fit_continuous, fit_indicator)
```

**Interpretation:** A small p-value suggests the more complex model (with categorical intensity) provides significantly better fit.

{{< pagebreak >}}

### Overall Model F-test
To test the overall significance of a model against the intercept-only model:

```{r}
# Fit full model
fit_full <- lm(Flowers ~ Intensity + Time, data = case0901)

# ANOVA table for overall significance
anova(fit_full)
```

**Key Components:**

- **F-statistic:** Ratio of mean square regression to mean square residual
- **p-value:** Probability of observing this F-statistic if null hypothesis (no relationship) is true

```{r}
# Manual F-statistic calculation
F_stat <- (1733.35 / 41.49)
p_value <- pf(F_stat, 2, 21, lower.tail = FALSE)
cat("F-statistic:", F_stat, "\np-value:", p_value)
```

{{< pagebreak >}}

## Confidence & Prediction Intervals

### Model Setup
We'll use the following model to predict log income:

```{r}
# Load and prepare data
data(ex0923)
head(ex0923)

# Fit model with na handling
income_model <- lm(log(Income2005) ~ AFQT + Educ + Gender,
                   data = ex0923,
                   na.action = na.exclude)
```

### Creating Prediction Grid
Generate combinations of predictor values for visualization:

```{r}
pred_grid <- expand.grid(
  Educ = unique(ex0923$Educ),
  AFQT = seq(0, 100, 10),
  Gender = unique(ex0923$Gender)
)
head(pred_grid)
```

### Making Predictions
Calculate estimated means and intervals:

```{r}
# Confidence intervals for mean response
conf_intervals <- predict(income_model, 
                         newdata = pred_grid,
                         interval = "confidence")

# Prediction intervals for individual responses
pred_intervals <- predict(income_model,
                         newdata = pred_grid,
                         interval = "prediction")

# Combine results
results <- cbind(pred_grid,
                conf_mean = conf_intervals[, 1],
                conf_lwr = conf_intervals[, 2],
                conf_upr = conf_intervals[, 3],
                pred_lwr = pred_intervals[, 2],
                pred_upr = pred_intervals[, 3])
head(results)
```

{{< pagebreak >}}

### Visualization Examples

**1. Mean Income by Education Level:**
```{r}
#| fig-height: 5
#| fig-width: 7
library(ggplot2)

ggplot(subset(results, Educ == 12), 
       aes(AFQT, conf_mean, color = Gender)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf_lwr, ymax = conf_upr), 
              alpha = 0.2) +
  labs(title = "Estimated Mean Income (12 Years Education)",
       y = "Log Income",
       x = "AFQT Score") +
  theme_minimal()
```

**2. Prediction Intervals Comparison:**
```{r}
#| fig-height: 5
#| fig-width: 7
ggplot(subset(results, Educ == 12 & Gender == "male"), 
       aes(AFQT, conf_mean)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf_lwr, ymax = conf_upr),
              alpha = 0.3, fill = "blue") +
  geom_ribbon(aes(ymin = pred_lwr, ymax = pred_upr),
              alpha = 0.2, fill = "red") +
  labs(title = "Male Income Prediction (12 Years Education)",
       subtitle = "Blue: 95% CI for Mean | Red: 95% Prediction Interval",
       y = "Log Income",
       x = "AFQT Score") +
  theme_minimal()
```

{{< pagebreak >}}

## Exercises


1. **Model Comparison:** Fit a model adding an `Educ:AFQT` interaction term. Compare it to our original model using an F-test. Interpret the results.

2. **Visualization:** Recreate the first plot with back-transformed values (original income scale instead of log scale). What changes in the interpretation?

3. **Interval Interpretation:** For a female with 16 years education and AFQT=85:
   - Calculate both confidence and prediction intervals
   - Explain the difference between these intervals in context

{{< pagebreak >}}

## Summary Table

| Concept               | Purpose                                  | R Function          |
|-----------------------|------------------------------------------|---------------------|
| Nested Model F-test   | Compare model complexity                 | `anova(model1, model2)` |
| Confidence Interval   | Estimate precision of mean response     | `predict(..., interval = "confidence")` |
| Prediction Interval   | Range for individual observations       | `predict(..., interval = "prediction")` |

**Key Takeaways:**

- F-tests help determine if additional predictors significantly improve model fit
- Confidence intervals quantify uncertainty about _mean_ responses
- Prediction intervals account for both model uncertainty and individual variation
