---
title: "ST565 Homework 1"
subtitle: "ST565: Time Series"
author: "Brian Cervantes Alvarez"
date: "01-31-2023"
format: OSULatexTheme-pdf
execute: 
  warning: false
  message: false
editor: visual
---

# Question 1

## Part A

### Finding $E(\bar{x})$
- $E(\bar{x}) = E\left(\frac{x_1 + x_2}{2}\right)$
- This becomes $\frac{E(x_1) + E(x_2)}{2}$.
- Given $E(x_1) = E(x_2) = \mu$, it simplifies to $E(\bar{x}) = \frac{\mu + \mu}{2} = \mu$.

### Finding $\text{Var}(\bar{x})$

- $\text{Var}(\bar{x}) = \text{Var}\left(\frac{x_1 + x_2}{2}\right)$
- The variance becomes $\frac{1}{4}\text{Var}(x_1 + x_2)$
- Since $\text{Var}(x_1 + x_2) = \text{Var}(x_1) + \text{Var}(x_2) + 2\text{Cov}(x_1, x_2)$ and $\text{Cov}(x_1, x_2) = 0$ because $\text{cor}(x_1, x_2) = 0$, it simplifies to $\frac{1}{4}(2\sigma^2)$ because $\text{Var}(x_1) = \text{Var}(x_2) = \sigma^2$.
- Thus, $\text{Var}(\bar{x}) = \frac{\sigma^2}{2}$.

## Part B

### Finding $E(\bar{x})$
- $E(\bar{x})$ does not depend on the correlation; hence, $E(\bar{x}) = \mu$.

### Finding $\text{Var}(\bar{x})$

- $\text{Var}(\bar{x}) = \frac{1}{4}\text{Var}(x_1 + x_2)$ still holds.
- Now, $\text{Var}(x_1 + x_2) = \text{Var}(x_1) + \text{Var}(x_2) + 2\text{Cov}(x_1, x_2)$, but $\text{Cov}(x_1, x_2) \neq 0$.
- Given $\text{Cov}(x_1, x_2) = \rho\sigma^2$ because $\text{cor}(x_1, x_2) = \rho$ and $\text{Cov}(x_1, x_2) = \text{cor}(x_1, x_2) \cdot \sigma_{x_1} \cdot \sigma_{x_2}$, the variance becomes $\frac{1}{4}(2\sigma^2 + 2\rho\sigma^2)$.
- Therefore, $\text{Var}(\bar{x}) = \frac{\sigma^2}{2}(1 + \rho)$.

## Part C

### Effect of Correlation on Variance of Sample Mean
- When $\text{cor}(x_1, x_2) = 0$, the variance of the sample mean is $\frac{\sigma^2}{2}$.
- When $\text{cor}(x_1, x_2) = \rho$, the variance of the sample mean becomes $\frac{\sigma^2}{2}(1 + \rho)$.
- The correlation affects the variance of the sample mean. Positive correlation increases the variance, whereas negative correlation decreases it.
- Ignoring correlation in inference can lead to underestimating or overestimating the variability in the data.

{{< pagebreak >}}

# Question 2

## Part A

### Show $n = 3, 4$
- For $n = 3, 4$, can be described by the formula $x_t = \delta t + \sum_{j=1}^{t} w_j$ for $t = 1, \ldots, n$.
- Let's show it for $t = 1, 2, 3, 4$:
    - When $t = 1$, $x_1 = \delta + x_0 + w_1 = \delta + w_1$.
    - When $t = 2$, $x_2 = \delta + x_1 + w_2 = \delta + (\delta + w_1) + w_2 = 2\delta + w_1 + w_2$.
    - When $t = 3$, $x_3 = \delta + x_2 + w_3 = \delta + (2\delta + w_1 + w_2) + w_3 = 3\delta + w_1 + w_2 + w_3$.
    - When $t = 4$, $x_4 = \delta + x_3 + w_4 = \delta + (3\delta + w_1 + w_2 + w_3) + w_4 = 4\delta + w_1 + w_2 + w_3 + w_4$.

## Part B

### Determining the Mean
- $E[x_t] = E[\delta + x_{t-1} + w_t]$.
- Since $E[w_t] = 0$ for white noise, $E[x_t] = \delta + E[x_{t-1}]$.
- If you continue the pattern, the mean becomes $\delta t$.

### Determining the Variance
- $\text{Var}(x_t) = \text{Var}(\delta + x_{t-1} + wt)$.
- Since $\text{Var}(w_t) = \sigma_w^2$ and $w_t$ is independent, the variance converges to $t\sigma_w^2$.

### Determining Stationarity
- The process is not stationary as the mean and variance depend on time $t$.

## Part C

- Autocovariance function is given as $\gamma(s, t)$ of $x_t$ for $\delta = 0$ where $\gamma(s, t) = \text{Cov}(x_s, x_t)$.
- For $\delta = 0$, $x_t = \sum_{j=1}^{t} w_j$.
- $\gamma(s, t)$ depends on the overlap of the sums of $w_j$. This means that if $s \leq t$, then the ACF is $\gamma(s, t) = s\sigma_w^2$.

## Part D

- The autocovariance function does the following compared to the random walk and moving average processes:
    - In the random walk process, the autocovariance function increases with time lag.
    - In a moving average process, the autocovariance function usually becomes zero after a certain lag, indicating no correlation beyond a specific period.
    
    
{{< pagebreak >}}
    
# Question 3

## Part A: $x_t = w_t - w_{t-3}$

### Stationarity
- The process is stationary. 
- Each $w_t$ is independent and identically distributed, and the structure of $x_t$ does not change over time.

### Mean
- The mean is $E[x_t] = E[w_t - w_{t-3}] = E[w_t] - E[w_{t-3}] = 0 - 0 = 0$.

### Autocovariance Function
- The autocovariance function depends on the lag $k$: 
    - For $k = 0$, $\text{Cov}(x_t, x_t) = \text{Var}(w_t - w_{t-3}) = \text{Var}(w_t) + \text{Var}(w_{t-3}) = 2$.
    - For $k = 3$ or $k = -3$, $\text{Cov}(x_t, x_{t-3}) = -1$.
    - Otherwise, it is $0$.

## Part B: $x_t = w_1^3$

### Stationarity
- The process is not stationary.
- $x_t$ is constant over time, which violates the requirement for stationarity that the mean and variance must not depend on time.

## Part C: $x_t = t + w_1^3$

### Stationarity
- The process is not stationary.
- The term $t$ introduces a time-dependent trend.

## Part D: $x_t = w_t^2$

### Stationarity
- The process is stationary.
- Each $w_t^2$ follows a Chi-squared distribution with 1 degree of freedom and is independent of time $t$.

### Mean
- The mean is $E[w_t^2] = 1$ (since the variance of $w_t$ is $1$).

### Autocovariance Function
- The autocovariance function depends on the lag $k$:
    - For $k = 0$, it is the variance of $w_t^2$, which is $2$.
    - For $k \neq 0$, it is $0$ since $w_t$ are independent.

## Part E: $x_t = w_t w_{t-2}$

### Stationarity
- The process is stationary.
- The multiplication of two independent, identically distributed normal random variables at different times does not depend on time $t$.

### Mean
- The mean is $E[w_t w_{t-2}] = E[w_t] \cdot E[w_{t-2}] = 0$.

### Autocovariance Function
- The autocovariance function depends on the lag $k$:
    - For $k = 2$ or $k = -2$, $\text{Cov}(x_t, x_{t-2}) = 1$.
    - Otherwise, it is $0$.


{{< pagebreak >}}

# Question 4

## Part A

```{r}
#| echo: false

library(extrafont)

extrafont::loadfonts(quiet = TRUE)

# Custom Theme Function
theme_osu <- function(base_size = 12, base_family = "Georgia") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      # Color and fill elements with OSU's colors
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "white", color = NA),
      plot.title = element_text(color = "#DC4405", size = rel(1.2)),
      plot.subtitle = element_text(color = "#DC4405", margin = margin(b = 10)),
      plot.caption = element_text(color = "black"),
      axis.title = element_text(color = "black"),
      axis.text = element_text(color = "black"),
      legend.text = element_text(color = "black"),
      legend.title = element_text(color = "black")
      # Add any other elements you wish to customize
    )
}
```




```{r}
#| echo: false

library(tidyverse)
library(lubridate)

leagueDs <- read.csv("leagueWorlds.csv", header = FALSE)
anitaDs <- read.csv("anitamaxwynn.csv", header = FALSE)
palworldDs <- read.csv("palworld.csv", header = FALSE)


# Remove incorrect rows

leagueDs <- leagueDs[-c(1,2),]
anitaDs <- anitaDs[-c(1,2),]
palworldDs <- palworldDs[-c(1,2),]

# Fix Column Headers

leagueDs <- leagueDs %>% 
  mutate(YearMonth = ym(V1),
         SearchCount = as.numeric(ifelse(V2 %in% "<1", 0, V2))) %>%
  select(-c(V1, V2))

anitaDs <- anitaDs %>% 
  mutate(Month = ymd(V1),
         SearchCount = as.numeric(ifelse(V2 %in% "<1", 0, V2))) %>%
  select(-c(V1, V2))

palworldDs <- palworldDs %>% 
  mutate(Day = ymd(V1),
         SearchCount = as.numeric(ifelse(V2 %in% "<1", 0, V2))) %>%
  select(-c(V1, V2))


#head(leagueDs)
#head(anitaDs)
#head(palworldDs)
```

```{r}
#| echo: false

p1 <- leagueDs %>% 
  ggplot(aes(x = YearMonth, y = SearchCount)) +
  geom_line() +
  labs(title = "Google Trends - Search Term: 'League of Legends Worlds'",
       subtitle = "From Jan 1st, 2011 to Feb 1st, 2021", 
       x = NULL,
       y = "Search Popularity (Scale 0-100)") +
  theme_osu()

p2 <- anitaDs %>% 
  ggplot(aes(x = Month, y = SearchCount)) +
  geom_line() +
  labs(title = "Google Trends - Search Term: 'Anita Max Wynn'",
       subtitle = "From Dec 17th, 2011 to Jan 28th, 2024", 
       x = NULL ,
       y = "Search Popularity (Scale 0-100)") +
  theme_osu()

p3 <- palworldDs %>% 
  ggplot(aes(x = Day, y = SearchCount)) +
  geom_line() +
  labs(title = "Google Trends - Search Term: 'Palworld'",
       subtitle = "From Jan 1st, 2024 to Jan 28th, 2024", 
       x = NULL,
       y = "Search Popularity (Scale 0-100)") +
  theme_osu()
```


#### Initial Popularity (2011-2013):
   - **Low Baseline Interest:** During this period, the interest in "League of Legends Worlds" was considerably low, reflecting poor interest early on. 
   
#### Growth Phase (2014 Onwards):
   - **Rapid Increase in Popularity:** This could be attributed to several factors, including the rising popularity of esports, increased marketing efforts by Riot Games, and the event's growing reputation as a premier competitive gaming spectacle.
   - **Annual Peaks:** These peaks are likely more pronounced and have a higher amplitude compared to earlier years, indicating heightened anticipation and engagement from the global gaming community.

```{r}
#| echo: false
p1 + scale_x_date(date_labels = "%Y", 
                  date_breaks = "2 year")
```


{{< pagebreak >}}

## Part B

- **Here's Some Context:**  Palworld, a video game blending elements from Ark: Survival Evolved, Legend of Zelda: Breath of the Wild, and incorporating Pokémon-like mechanics, saw exponential growth in search interest post its early access launch starting on January 19th 2024, showcasing a dramatic spike in public attention.

- **Potential Cause in Popularity:** Top streamers from Twitch began to stream Palworld, causing a major influx of gamer attention. Even a few of my friends started playing it. In other words, they got on board of what I like to call the "current streamer game train". 

- **Where it is trending?** After the initial surge, the game stabilized and then entered a phase of steady linear growth from January 22, 2024, onwards, reflecting a sustained interest in the game's unique gameplay experience. Maybe it's a good Pokémon-like game?


```{r}
#| echo: false
p3 + scale_x_date(date_labels = "%b %d", 
                  date_breaks = "week")
```

{{< pagebreak >}}

## Part C

- **Event-Driven Spike:** A sharp increase in search interest for "Anita Max Wynn" was triggered by a mention from rapper Drake, peaking sharply on December 25th due to viral spread and memeing.
- **Rapid Decline and Leveling Off:** After peaking, interest in the term quickly declined from late January, showing the nature of viral meme culture, yet it leveled off slightly above pre-spike levels, indicating some residual interest-probably due to streamers referencing that meme joke.
- **Absence of Seasonality and Trend:** It does not follow predictable seasonal patterns or a long-term trend, highlighting its dependence on being a viral meme.

```{r}
#| echo: false
p2 + scale_x_date(date_labels = "%b %d", 
                  date_breaks = "week")
```

