---
title: "ST 352 | Lab Assignment 3 - Guide"
subtitle: "Regression Analysis and Inference"
author: "Brian Cervantes Alvarez"
date: "10-15-2024"
format: pdf
execute: 
  echo: true
  warning: false
webr:
  packages: ['ggplot2', 'readr', 'dplyr', 'tidyr']
  show-startup-message: false
filters:
- webr
editor:
    render-on-save: true
---

**Reminder of the honor code:**  
*Lab assignments are to be completed individually!*

# Objective

In this lab assignment, you will explore regression analysis and inference using the dataset `reaction.txt`. Problem 1 focuses on investigating the relationship between age and reaction time.

# Problem 1: Age and Reaction Time

During Week 3, you participated in an activity to test your reaction time. For this problem, you will investigate the relationship between age and reaction time. Specifically, you aim to determine if there is evidence to suggest that a person's age can predict their reaction time.

The data for this analysis are stored in the `reaction.txt` dataset on Canvas. The dataset contains two variables:

- `age` (in years)
- `time` (average of three reaction times in milliseconds)

{{< pagebreak >}}

## Part I | Assessing Conditions

### 1) 

**Hint**

Generate three plots: a scatterplot of reaction time versus age, a residual plot to assess residuals, and a Normal Q-Q plot to evaluate the normality of residuals.



```{r}
# Load the dataset
reactionData <- read.table("reaction.txt", header = TRUE)

# 1. Scatterplot (Base R)
plot(reactionData$age, reactionData$time, 
     main = "Scatterplot of Reaction Time vs. Age",
     xlab = "Age (years)", ylab = "Reaction Time (ms)", pch = 19, col = "blue")


# Fit the linear regression model
modelReaction <- lm(time ~ age, data = reactionData)

# Calculate residuals
reactionData$residuals <- residuals(modelReaction)

# 2. Residual Plot (Base R)
plot(reactionData$age, reactionData$residuals, 
     main = "Residual Plot", 
     xlab = "Age (years)", ylab = "Residuals (ms)", pch = 19, col = "red")
abline(h = 0, lty = 2)

# 3. Normal Q-Q Plot (Base R)
qqnorm(reactionData$residuals, main = "Normal Q-Q Plot of Residuals",
                               pch = 19, col = "darkgreen")
qqline(reactionData$residuals)
```

### 2) Are there any outliers or other deviations from the pattern? Which plot(s) are you using to assess the outlier condition?


**Hint**

Examine the **Residual Plot** and the **Scatterplot** to identify any data points that fall far from the general trend. Look for points that are noticeably distant from the main cluster of data.


### 3) Is the linearity condition satisfied? Explain. In your explanation, reference the plot(s) you are using to assess the linearity condition.


**Hint**

Consider the overall trend in the **Scatterplot**. Check if the relationship between age and reaction time appears to follow a straight line without any noticeable curves.


### 4) Is the constant variation condition satisfied? Explain. In your explanation, reference the plot(s) you are using to assess the constant variation condition.


**Hint**

Look at the spread of residuals in the **Residual Plot**. Determine if the variability of residuals remains consistent across all levels of age.


### 5) Is the normality condition satisfied? Explain. In your explanation, reference the plot(s) you are using to assess the normality condition.


**Hint**

Review the **Normal Q-Q Plot** of the residuals. Assess whether the points approximately follow the reference line, indicating normal distribution of residuals.


### 6) Is a transformation necessary in this example? Explain.


**Hint**

Think about whether the assumptions of linear regression are adequately met. If all conditions (linearity, constant variation, normality) are satisfied, a transformation might not be needed.

{{< pagebreak >}}

## Part II | Inference in Regression

### 7) Write the least-squares regression equation. Define the terms in the equation in the context of the problem.


**Hint**

The regression equation relates the predicted reaction time to age. Identify the intercept and slope from the linear model.


### 8) Is there evidence that a person’s age can be used to predict their reaction time?

#### a) State the null and alternative hypotheses in notation and words.


**Hint**

Formulate the hypotheses to test whether the slope of age is significantly different from zero.


#### b) From the regression output, report the appropriate test-statistic with degrees of freedom.


**Hint**

Identify the t-statistic and the corresponding degrees of freedom from the summary of the regression model.


#### c) State a conclusion in the context of the problem. Support your conclusion with a p-value.


**Hint**

Based on the p-value obtained from the regression output, decide whether to reject or fail to reject the null hypothesis.


### 9) Use R to predict the reaction time of a 22-year-old. Report your answer here.

```{r}
#| eval: false
# Predict reaction time for a 30-year-old [Example]
newPerson <- data.frame(age = 30) 
predictedTime <- ...(..., newdata = ...)
predictedTime
```


**Hint**

Use the `predict()` function with the fitted model and a new data frame containing the age value of 22.

### 10) Obtain the appropriate 95% interval. Report and interpret the appropriate interval.

#### a) A student wants to estimate the average reaction time for all 22-year-old people.

```{r}
#| eval: false
# Confidence interval for the mean reaction time at age 30 [Example]
confIntervalMean <- ...(..., newdata = ..., ..., ...)
confIntervalMean
```


**Hint**

Use the `predict()` function with `interval = "confidence"` to obtain the confidence interval for the mean reaction time.


#### b) A student wants to estimate the reaction time for a person who is 22 years old.

```{r}
#| eval: false
# Prediction interval for an individual reaction time at age 30 [Example]
predInterval <- ...(..., newdata = ..., ...)
predInterval
```


**Hint**

Use the `predict()` function with `interval = "prediction"` to obtain the prediction interval for an individual reaction time.


{{< pagebreak >}}

# Appendix

## Want to do it better? Use ggplot2 & tidyverse packages!
```{r}
# Load tidyverse libraries
library(readr)
library(ggplot2)
library(dplyr)

# Read the reaction data
reactionData <- read_table("reaction.txt")

# Fit the linear regression model
modelReaction <- lm(time ~ age, data = reactionData)

# Calculate residuals
reactionData <- reactionData %>%
  mutate(residuals = resid(modelReaction))

# 1. Scatterplot
ggplot(reactionData, aes(x = age, y = time)) +
  geom_point(color = "blue") +
  labs(title = "Scatterplot of Reaction Time vs. Age",
       x = "Age (years)",
       y = "Reaction Time (ms)") +
  theme_minimal()

# 2. Residual Plot
ggplot(reactionData, aes(x = age, y = residuals)) +
  geom_point(color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot",
       x = "Age (years)",
       y = "Residuals (ms)") +
  theme_minimal()

# 3. Normal Q-Q Plot of the Residuals
ggplot(reactionData, aes(sample = residuals)) +
  stat_qq(color = "darkgreen") +
  stat_qq_line(color = "black") +
  labs(title = "Normal Q-Q Plot of Residuals") +
  theme_minimal()

```

## Confidence Interval vs. Prediction Interval
```{r}
library(broom)

# Use the augment function to add fitted values and intervals
augmentedData <- augment(modelReaction, newdata = reactionData, 
                                        interval = "confidence", se_fit = TRUE)

# Add prediction intervals
augmentedData <- augmentedData %>%
  mutate(pred.lower = predict(modelReaction, newdata = reactionData, 
                                             interval = "prediction")[, 2],
         pred.upper = predict(modelReaction, newdata = reactionData,
                                             interval = "prediction")[, 3])

# 1. Plot with Confidence Interval
ggplot(augmentedData, aes(x = age, y = time)) +
  geom_point(color = "blue") +
  geom_line(aes(y = .fitted), color = "black") +
  geom_ribbon(aes(ymin = .lower, ymax = .upper), fill = "blue", alpha = 0.2) +
  labs(title = "Reaction Time vs. Age with Confidence Interval",
       x = "Age (years)", y = "Reaction Time (ms)") +
  theme_minimal()

# 2. Plot with Prediction Interval
ggplot(augmentedData, aes(x = age, y = time)) +
  geom_point(color = "blue") +
  geom_line(aes(y = .fitted), color = "black") +
  geom_ribbon(aes(ymin = pred.lower, ymax = pred.upper), fill = "red", alpha = 0.2) +
  labs(title = "Reaction Time vs. Age with Prediction Interval",
       x = "Age (years)", y = "Reaction Time (ms)") +
  theme_minimal()

```
