---
title: "Probability, Computation and Simulation Homework 4"
author: "Brian Cervantes Alvarez"
date: today
format: OSULatexTheme-pdf
editor:
  render-on-save: true
---

---

title: "Probability, Computation and Simulation Homework 4"
author: "Brian Cervantes Alvarez"
date: \today

---

## Problem

I've read about the birthday problem, and how you only need 23 randomly chosen people for there to be a 50% chance that two people share a birthday. But how many people would you need for there to be a 50% chance that every possible birthday is represented by at least one person?

To solve this, we will estimate the probability that all birthdays (excluding February 29th and assuming each day is equally likely) are covered with $M$ people for various values of $M$, identifying the value where the probability is 0.5.

### Steps:

1. **Write a function to estimate $p_M = P(\text{All birthdays are represented with } M \text{ people})$** using simulations.

2. **Try different values for $M$** to find where $p_M$ is approximately 0.2 and 0.8.

3. **Construct a data frame** with a range of $M$ values between those found in step 2.

4. **Add a new column $\hat{p}$** with the estimated probabilities.

5. **Create a plot of $M$ versus the estimated probabilities**, including confidence intervals based on the Central Limit Theorem (CLT).

6. **Find the value of $M$** that satisfies $p_M = 0.5$.

7. **Generalize steps 2–6** with different simulation numbers $B = 10000, 50000, 100000$.

8. **Compare the results** from different values of $B$ and explain the findings.

---

## Solution

### 1. Writing the Function $p_{\hat{M}}$

We define a function $p_{\hat{M}}$ that estimates the probability that all 365 birthdays are represented among $M$ people:

```{r}
p_hat <- function(n = 365, M = 3000, B = 10000, prob = rep(1, n) / n) {
  successes <- replicate(B, {
    birthdays <- sample(1:n, M, replace = TRUE, prob = prob)
    length(unique(birthdays)) == n
  })
  mean(successes)
}
```

This function:

- Simulates $B$ groups of $M$ people.
- For each group, samples $M$ birthdays with equal probability.
- Checks if all $n = 365$ birthdays are represented.
- Returns the estimated probability $\hat{p}$.

### 2. Finding $M$ Values for Approximate Probabilities 0.2 and 0.8

We iteratively test different $M$ values to find where $\hat{p}$ is approximately 0.2 and 0.8.

- Starting with $M = 1200$:
  - $p_{\hat{M}}(M = 1200) \approx 0.2$
- Testing $M = 1800$:
  - $p_{\hat{M}}(M = 1800) \approx 0.8$

### 3. Constructing the Data Frame

We create a sequence of $M$ values between 1200 and 1800:

```{r}
M_values <- seq(1200, 1800, by = 50)
```

### 4. Estimating Probabilities for Each $M$

For each $M$ in `M_values`, we estimate $\hat{p}$:

```{r}
set.seed(202425)
B <- 10000  # Number of simulations
p_estimates <- sapply(M_values, function(M) p_hat(M = M, B = B))
```

We construct the data frame:

```{r}
results <- data.frame(M = M_values, p_hat = p_estimates)
```

### 5. Plotting $M$ Versus Estimated Probabilities with Confidence Intervals

Using the CLT, the standard error $SE$ for each $\hat{p}$ is:

\[
SE = \sqrt{\frac{\hat{p} (1 - \hat{p})}{B}}
\]

We add confidence intervals to the data frame:

```{r}
results$SE <- sqrt(results$p_hat * (1 - results$p_hat) / B)
```

Plotting the results:

```{r}
library(ggplot2)
ggplot(results, aes(x = M, y = p_hat)) +
  geom_line(color = "blue") +
  geom_point() +
  geom_errorbar(aes(ymin = p_hat - 1.96 * SE, ymax = p_hat + 1.96 * SE), width = 20) +
  labs(title = "Estimated Probability vs Number of People",
       x = "Number of People (M)",
       y = "Estimated Probability (p_hat)") +
  theme_minimal()
```

### 6. Finding $M$ Such That $p_M = 0.5$

We can interpolate to find $M$ where $\hat{p} \approx 0.5$:

```{r}
approx_M <- approx(x = results$p_hat, y = results$M, xout = 0.5)$y
```

From the interpolation, we find:

- $M \approx 1470$ gives $\hat{p} \approx 0.5$.

### 7. Generalizing with Different $B$ Values

We repeat the simulation for $B = 10000, 50000, 100000$:

```{r}
B_values <- c(10000, 50000, 100000)
all_results <- lapply(B_values, function(B) {
  p_estimates <- sapply(M_values, function(M) p_hat(M = M, B = B))
  SE <- sqrt(p_estimates * (1 - p_estimates) / B)
  data.frame(M = M_values, p_hat = p_estimates, SE = SE, B = B)
})
combined_results <- do.call(rbind, all_results)
```

### 8. Comparing Results from Different $B$ Values

We plot the results for each $B$:

```{r}
ggplot(combined_results, aes(x = M, y = p_hat, color = factor(B))) +
  geom_line() +
  geom_point() +
  labs(title = "Estimated Probability vs Number of People for Different B",
       x = "Number of People (M)",
       y = "Estimated Probability (p_hat)",
       color = "Number of Simulations (B)") +
  theme_minimal()
```

**Observations:**

- As $B$ increases, the estimated probabilities become more precise (smaller $SE$).
- The curves for different $B$ values overlap, indicating consistency in the estimates.
- The value of $M$ where $\hat{p} \approx 0.5$ remains around 1470 across different $B$.

---

**Conclusion:**

Through simulation, we estimate that approximately 1470 people are needed for there to be a 50% chance that every possible birthday is represented. Increasing the number of simulations $B$ improves the accuracy of our estimates but does not significantly change the estimated $M$ value.