---
title: "ST557: Homework 1"
author: 
    - "Brian Cervantes Alvarez"
date: "10-06-2023"
format: PrettyPDF-pdf
execute: 
  warning: false
  message: false
editor: visual
---

# Question 1

```{r}
# Set random seed for reproducibility
set.seed(503)

# Read HW1Q1 dataset in using base R
ds <- read.csv("HW1Q1.csv")
# Show first 5 rows
head(ds, 5)


```

## Part A

Read this data into R using read.csv(). Create a 2-dimensional scatter plot of the 20 observations (use plot() function in R).

```{r}
# Get X and Y Components into vectors
x = ds$X
y = ds$Y

# Create scatter plot of Y ~ X
plot(x,y)


pnorm(19.9, 20, sqrt(4/40))


pnorm(0.5, 20, sqrt(4/8))

```

{{< pagebreak >}}

## Part B

Find the sample mean vector, and add this point to the plot using points(). You can make this point a different color using the col= argument, or you can make it a different plotting character using the pch= argument. For example: \> points(sampMean\[1\], sampMean\[2\], pch=16, col=2

```{r}
plot(x,y)

sampMeanX = mean(x)
sampMeanY = mean(y)

points(sampMeanX, sampMeanY, pch=16, col=2)

```

{{< pagebreak >}}

## Part C

Find the sample covariance matrix.

```{r}
# Sample covariance matrix
sampCov <- cov(ds)

sampCov
```
{{< pagebreak >}}

## Part D

Find the eigendecomposition (spectral decomposition) of the sample covariance matrix using eigen().

```{r}
sampEigenDecomp <- eigen(sampCov)

sampEigenDecomp
```
{{< pagebreak >}}

## Part E

Add the eigenvector corresponding to the largest eigenvalue to the plot as a vector from the sample mean using `lines()`. Be careful here: if the first eigenvector is $(v_1, v_2)$ and the sample mean vector is $(\bar{x}_1, \bar{x}_2)$, you want a line from $(\bar{x}_1, \bar{x}_2)$ to $(\bar{x}_1 + v_1, \bar{x}_2 + v_2)$."

```{r}
# Find the index of the largest eigenvalue
largestEigenvalue <- which.max(sampEigenDecomp$values)
# Extract the eigenvector corresponding to the largest eigenvalue
largestEigenvector <- sampEigenDecomp$vectors[, largestEigenvalue]

# Calculate the endpoints of the line segment
sampMeanX <- mean(x)
sampMeanY <- mean(y)
xend <- sampMeanX + largestEigenvector[1]
yend <- sampMeanY + largestEigenvector[2]

plot(x,y)
# Add the line segment to the plot
lines(c(sampMeanX, xend), c(sampMeanY, yend), col = "seagreen4", lwd = 2)
points(sampMeanX, sampMeanY, pch=16, col=2)

```

### Discussion for Part E

**Question: Describe how the direction of this eigenvector relates to the cloud of data points**

The eigenvector follows the trend of the data points, which themselves are showing a positive increasing trend, and it culminates precisely at the location of the sample means for X and Y.

{{< pagebreak >}}

# Question 2

```{r}



```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
