---
title: "Homework Guide: Multifactor Studies & Time Series Analysis"
author: "TA Solutions"
date: today
format: html
---

# Introduction

This homework guide provides step-by-step instructions for two problems:

1. **Problem 1** (Multifactor studies without replication) using a small soybean dataset.
2. **Problem 2** (Adjustment for serial correlation) using an annual temperature dataset.

The goal is for students to learn how to set up, interpret, and communicate their statistical analyses.

---

# Problem 1: Multifactor Studies Without Replication

You have a dataset on seed yield of two soybean cultivars (Forrest and William) grown under different levels of:

- Soil Moisture Stress (categorical: "Well-watered" vs. other levels)
- SO₂ (continuous)
- O₃ (continuous)
- Yield for two cultivars (Forrest, William)

An example slice of the data:

| Stress        | SO2  | O3   | Forrest | William |
|---------------|------|------|---------|---------|
| Well-watered  | 0.00 | 0.02 | 4376    | 5561    |
| Well-watered  | 0.00 | 0.05 | 4544    | 5947    |
| Well-watered  | 0.00 | 0.07 | 2806    | 4273    |
| Well-watered  | 0.00 | 0.08 | 3339    | 3470    |
| Well-watered  | 0.00 | 0.10 | 3320    | 3080    |
| Well-watered  | 0.02 | 0.02 | 3747    | 5092    |

## (a) Effect of Soil Moisture Stress on Yield

We want to examine if **Soil Moisture Stress** affects yield beyond O₃ and SO₂.

1. **Reshape the data** into a long format so we have one "Yield" column:

```{r}
library(tidyr)
library(dplyr)

soy_data <- case1402

# Suppose your data frame is named `soy_data`.

tidy_data <- soy_data %>%
  pivot_longer(
    cols = c(Forrest, William),
    names_to = "Cultivar",
    values_to = "Yield"
  )

head(soy_data)
head(tidy_data)



```

2. **Fit an additive model**:

```{r}
model1 <- lm(Yield ~ Stress + SO2 + O3 + Cultivar, data = tidy_data)
summary(model1)
model2 <- lm(Yield ~ Stress + SO2 + O3, data = tidy_data)
summary(model2)


williamModel <- lm(William ~ Stress + SO2 + O3, data = soy_data)
summary(williamModel)
forrestModel <- lm(Forrest ~ Stress + SO2 + O3, data = soy_data)
summary(forrestModel)


```

- **Interpretation**:
  - Look at the coefficient and p-value for `Stress`. A small p-value (< 0.05) indicates that Stress significantly influences yield after adjusting for O₃, SO₂, and Cultivar.

## (b) Testing Interactive Effects

Next, check if there is any *interaction* among the factors (Stress, SO₂, O₃):

```{r}
interaction_model <- lm(Yield ~ Stress * SO2 * O3 + Cultivar, data = tidy_data)
summary(interaction_model)
```

- **Interpretation**:
  - If interaction terms (e.g., `Stress:O3`) are significant, the effect of one variable depends on another.
  - If no interactions are significant, the additive model might be sufficient.

---

# Problem 2: Adjustment for Serial Correlation

You have annual temperature anomaly data for the Southern Hemisphere with columns:

| Year | Temperature |
|------|------------|
| 1850 | -0.499     |
| 1851 | -0.383     |
| 1852 | -0.431     |
| 1853 | -0.452     |
| 1854 | -0.478     |
| 1855 | -0.436     |

We want to model Temperature over time and check for possible autocorrelation.

## (a) Quadratic Regression and Residual Plot

1. **Fit the quadratic model**:

```{r}
model_quad <- lm(Temperature ~ Year + I(Year^2), data = temp_data)
summary(model_quad)
```

2. **Plot the residuals**:

```{r}
res <- resid(model_quad)
plot(temp_data$Year, res, type = "o", main = "Residuals vs. Year")
abline(h = 0, col = "red")
```

- **Interpretation**: Look for any patterns or runs of positive/negative residuals.

## (b) Partial Autocorrelation Plot

Check if the residuals have autocorrelation:

```{r}
# Generate PACF
pacf(res, main = "Partial ACF of Residuals")
```

- **Interpretation**: A spike at lag 1 suggests AR(1) autocorrelation.

## (c) Estimate the First Serial Correlation Coefficient (\(\alpha\))

```{r}
alpha_hat <- cor(res[-1], res[-length(res)])
alpha_hat
```

- **Interpretation**: \(\alpha\) is between -1 and 1.

## (d) Cochrane-Orcutt Procedure

Adjust for autocorrelation by iteratively estimating \(\rho\) and refitting:

```{r}
library(orcutt)
co_model <- cochrane.orcutt(model_quad)
summary(co_model)
```

- **Interpretation**: New estimates of coefficients and an estimate of \(\rho\) that accounts for serial correlation.

## (e) 95% Confidence Intervals (Manual Calculation)

If you need manual CIs:

$$
CI_{95\%}(\beta_j) = \hat{\beta}_j \pm t_{\alpha/2,df}\times SE(\hat{\beta}_j).
$$

- Use degrees of freedom \(n - k\) or adjusted for AR(1) fits.

---

# Final Teaching Tips

1. **Factorial Designs** (Problem 1): Highlight how to interpret main effects and interactions.
2. **Time Series** (Problem 2): Emphasize the importance of diagnostic checks (residual plots, PACF) when data are time-ordered.
3. **Interpretation**:
   - Encourage students to frame findings in plain language.
   - Example: "Moisture Stress significantly decreases yield" or "Temperature anomalies exhibit autocorrelation."

