---
title: "Lab 3: ST 512"
subtitle: "Inference on Multiple Linear Regression in R"
format: html
editor: visual
---

# Goals for Today’s Lab

- Perform hypothesis tests.
- Construct confidence intervals for single parameters.
- Answer questions related to practice quiz examples.

---

## Introduction

This lab uses the `case1002` dataset to explore inference in multiple linear regression. Key areas include:

- Testing the coefficient on the echo-locating bat indicator variable.
- Re-leveling categorical variables for meaningful interpretation.

---

## Step 1: Fitting the Model

```{r}
# Load necessary libraries
library(Sleuth3)
library(ggplot2)

# Fit the regression model
fit_bats <- lm(log(Energy) ~ log(Mass) + Type, data = case1002)
summary(fit_bats)
```

### Key Output:

- Coefficients provide estimates, standard errors, t-values, and p-values.
- Residual standard error and adjusted R-squared assess model fit.

### Extracting Coefficients:

```{r}
coefficients(fit_bats)
summary(fit_bats)$coefficients
summary(fit_bats)$sigma
```

---

## Step 2: Changing Reference Levels

### Rationale:
The default reference level (`echolocating bats`) does not align with the class model. Modify the reference level to `non-echolocating bats` for direct comparison.

```{r}
# Re-level Type to use non-echolocating bats as reference
case1002$Type2 <- relevel(case1002$Type, ref = "non-echolocating bats")

# Fit the new model
fit_2 <- lm(log(Energy) ~ log(Mass) + Type2, data = case1002)
summary(fit_2)
```

---

## Exercises

### 1. Comparing Non-Echolocating Bats and Birds

#### Steps:

1. Write the model, null, and alternate hypotheses:
   - Model: $\mu(\log(Energy) | \log(Mass), Type) = \beta_0 + \beta_1 \log(Mass) + \beta_2\text{ebat} + \beta_3\text{bird}$
   - $H_0: \beta_3 = 0$, $H_1: \beta_3 \neq 0$.

2. Fit the model:

```{r}
case1002$ebat <- ifelse(case1002$Type == "echolocating bats", 1, 0)
case1002$bird <- ifelse(case1002$Type == "non-echolocating birds", 1, 0)
fit_3 <- lm(log(Energy) ~ log(Mass) + ebat + bird, data = case1002)
summary(fit_3)
```

3. State the t-statistic and p-value.
4. Decision: Compare the p-value to $\alpha = 0.05$.

### 2. Comparing Echolocating Bats and Non-Echolocating Birds

Follow the same steps as above but adjust the reference category for interpretation.

### 3. Testing Differences in Slopes

Fit a model to test if the slope between `log(Mass)` and `log(Energy)` differs across groups:

- Null hypothesis: Slopes are the same.
- Alternate hypothesis: Slopes differ.

---

## Confidence Intervals

### Confidence Interval for Single Parameter

```{r}
confint(fit_2)
```

### Manual Calculation:

Use the formula from class to validate results:

```{r}
-1.57636 + c(-1, 1) * qt(0.975, 16) * 0.28724
```

---

## Practice Quiz Questions

**Question:** Test if the mean of `log(Energy)` differs across groups (e.g., bats vs. birds).

1. State the null and alternate hypotheses.
2. Fit the model.
3. Calculate t-statistic and p-value.
4. Interpret results.

---

## Functions Summary

### `lm()`
- **Purpose:** Fits linear models.
- **Common Parameters:**
  - `formula`: Specifies the model (e.g., `y ~ x1 + x2`).
  - `data`: Specifies the dataset.
- **When to Use:** Use this function when you want to model relationships between predictors and a response variable.

### `summary()`
- **Purpose:** Summarizes the results of a model.
- **Common Parameters:**
  - Applied to a model object (e.g., `summary(fit)`).
- **When to Use:** Use this function to view estimates, significance, and model diagnostics.

### `relevel()`
- **Purpose:** Changes the reference level of a factor.
- **Common Parameters:**
  - `x`: The factor variable.
  - `ref`: The new reference level.
- **When to Use:** Use this when you need a different reference level for interpreting coefficients.

### `confint()`
- **Purpose:** Calculates confidence intervals for model parameters.
- **Common Parameters:**
  - Applied to a model object (e.g., `confint(fit)`).
- **When to Use:** Use this to understand the range within which parameter estimates fall with a specified level of confidence.

### `ifelse()`
- **Purpose:** Creates a binary variable based on a condition.
- **Common Parameters:**
  - `test`: Logical condition.
  - `yes`: Value if the condition is `TRUE`.
  - `no`: Value if the condition is `FALSE`.
- **When to Use:** Use this when creating indicator variables or categorizing data.

---

