---
title: "Homework 3 Solutions Guide"
author: "Brian Cervantes Alvarez"
date: "2023-10-15"
format: pdf
---

# Problem 1: Brain Weights and Extra Sum of Squares F-Test

**Question:** After accounting for body weight, do gestation period and litter size affect brain weight?  

## Step-by-Step Explanation  

1. **What are we testing?**  
   We want to know if two variables (**gestation period** and **litter size**) have a relationship with brain weight *after* we already include body weight in our model.  

2. **How do we test this?**  
   - **Full Model:** Includes body weight, gestation, and litter size.  
   - **Reduced Model:** Only includes body weight.  
   We compare these two models using an **F-test**.  

3. **What does the F-test tell us?**  
   - If adding gestation and litter size significantly improves the model.  
   - **Null Hypothesis (H₀):** Neither gestation nor litter size matters.  
   - **Alternative Hypothesis (Hₐ):** At least one of them matters.  

4. **Results from the ANOVA Table:**  
   - **F-statistic:** 23.68  
   - **p-value:** < 0.0001 (extremely small!)  
   - **Conclusion:** The p-value is less than 0.05, so we reject H₀. This means gestation period, litter size, or both are significantly associated with brain weight.  

5. **Why log-transform the variables?**  
   Body and brain weights are often skewed. Taking the log (lBrain, lBody) helps make the relationship linear and easier to model.  

```{r}
# Corrected Code (Original had a typo with "-" instead of "~")
library(Sleuth3)
data(case0902)
Data <- case0902

# Log-transform variables
Data$lBrain <- log(Data$Brain)
Data$lBody <- log(Data$Body)
Data$llitter <- log(Data$litter)
Data$lGestation <- log(Data$Gestation)

# Full and reduced models
lmod_full <- lm(lBrain ~ lBody + lGestation + llitter, data = Data)
lmod_reduced <- lm(lBrain ~ lBody, data = Data)

anova(lmod_reduced, lmod_full) # Compare models
```

# Problem 2: Butterfly Occurrences  

## Part (a)  
**Question:** Does reserve size affect the number of butterfly species after accounting for observation days?  

### Key Results:  
- **Two-sided p-value:** 0.2443  
- **One-sided p-value (for positive effect):** 0.1222  

### Explanation:  
- The two-sided p-value is **not significant** (> 0.05), meaning we *fail to reject* the null hypothesis.  
- If we test for a **positive effect** (one-sided), the p-value is still 0.12, which is not strong enough evidence.  
- **Takeaway:** There’s no strong evidence that reserve size impacts species count *after accounting for observation days*.  

⚠️ **Why does observation time matter?**  
Researchers spent more days searching in larger reserves. More time = more species found, which could make reserve size *seem* important when it’s actually the effort!  

{{< pagebreak >}}

## Part (b)  
**Question:** What’s the p-value for testing if the coefficient of log(size) is 1?  

### Calculation:  
1. **t-statistic:**  
   $$
   t = \frac{0.0809 - 1}{0.1131} = -8.126
   $$  
2. **p-value:**  
   Very small ($1.88 \times 10^{-6}$).  

**Conclusion:** The coefficient is *significantly different from 1*.  

```{r}
# Code for part (b)
t_ratio <- (0.0809 - 1) / 0.1131
p_value <- 2 * (1 - pt(abs(t_ratio), df = 13))
p_value # Output: 1.88e-06
```

{{< pagebreak >}}

## Part (c)  
**Question:** What’s the 95% confidence interval for the coefficient of log(size)?  

### Calculation:  
- **Formula:**  
  $$
  0.0809 \pm (2.16 \times 0.1131) = (-0.163, 0.325)
  $$  
- **Interpretation:** The interval includes 0, meaning the effect of reserve size could be zero (supports part (a)).  

{{< pagebreak >}}

## Part (d)  
**Question:** How much variation in species count is *unexplained* by reserve size and observation days?  

### Answer:  
- **R² = 11.41%** (explained variation).  
- **Unexplained:** 100% - 11.41% = **88.59%**.  

**Translation:** Even with reserve size and observation days, most of the variation in butterfly species remains a mystery! Other factors (like habitat quality) might matter.  
