---
title: "ST 352 | Lab Assignment 1 SOLUTIONS"
subtitle: "Lab Demonstration"
author: "Brian Cervantes Alvarez"
date: "08-12-2024"
format: 
  html:
    toc: true
    toc-location: right
    toc-depth: 5
    html-math-method: katex
    page-layout: article
execute: 
  echo: true
  warning: false
webr:
  packages: ['mosaic', 'ggplot2', 'readr', 'dplyr', 'tidyr']
  show-startup-message: false
filters:
- webr
---


# Reminder of the Honor Code

Lab assignments are to be completed individually. Number your answers, put questions in order, and upload your assignment to Gradescope.

## Example

Is there evidence that OSU students watch more than 90 minutes of TV per day on average? The dataset `television.csv` contains the number of minutes per day for all participants.

### Question 1

**What is the variable of interest? Is it categorical or quantitative?**

- Time spent watching TV per day (in minutes). This is a **quantitative** variable.

### Question 2

**Is the sample representative of the population of OSU students? Why or why not?**

Answers may vary but should discuss whether the sample was randomly drawn. The TV-watching habits of ST 352 students may not be representative of the general OSU population.

# Exploring the Sample Data

### Question 3

**Summary Statistics**

```{r}
# Load necessary libraries
library(readr)
install.packages("readr")

tvDs <- read_csv("Academic-Year-2024-2025/Fall 2024/ST-352-Labs/Lab1/television.csv") 
tvDs2 <- read.csv("Academic-Year-2024-2025/Fall 2024/ST-352-Labs/Lab1/television.csv")

head(tvDs)
max(tvDs)

tv_data <- tvDs
tv_data <- tv_data[-70, ]


# Calculate the statistics
sample_size <- length(tv_data$minutes)
minimum <- min(tv_data$minutes, na.rm = TRUE)
first_quartile <- quantile(tv_data$minutes, 0.25, na.rm = TRUE)
median_value <- median(tv_data$minutes, na.rm = TRUE)
mean_value <- mean(tv_data$minutes, na.rm = TRUE)
third_quartile <- quantile(tv_data$minutes, 0.75, na.rm = TRUE)
maximum <- max(tv_data$minutes, na.rm = TRUE)
standard_deviation <- sd(tv_data$minutes, na.rm = TRUE)

# Print results
print(paste("Sample size:", sample_size))
print(paste("Minimum:", minimum, "minutes"))
print(paste("1st Quartile:", first_quartile, "minutes"))
print(paste("Median:", median_value, "minutes"))
print(paste("Mean:", mean_value, "minutes"))
print(paste("3rd Quartile:", third_quartile, "minutes"))
print(paste("Maximum:", maximum, "minutes"))
print(paste("Standard deviation:", standard_deviation, "minutes"))
```

| Statistic             | Value             |
|-----------------------|-------------------|
| Sample size           | 101               |
| Minimum               | 0 minutes         |
| 1st quartile          | 30 minutes        |
| Median                | 60 minutes        |
| Mean                  | 76.80 minutes     |
| 3rd quartile          | 90 minutes        |
| Maximum               | 1360 minutes      |
| Standard deviation    | 141.10 minutes    |

---

### Question 4

**Include your histogram and normal probability plot:**

```{r}
library(ggplot2)

# Histogram
ggplot(tv_data, aes(x = minutes)) + 
  geom_histogram(binwidth = 30) + 
  labs(title = "TV Minutes per Day", x = "Minutes", y = "Count") +
  theme_minimal()

# Normal Probability Plot
ggplot(tv_data, aes(sample = minutes)) + 
  stat_qq() + 
  stat_qq_line() + 
  labs(title = "Normal Q-Q Plot") +
  theme_minimal()
```

---

# Inference Using t-Methods

### Question 5

**Can t-methods be used for inference?**

Yes, due to the large sample size, t-methods are appropriate, even if the sample data appears skewed.

---

### Question 6

**Do you think the null hypothesis will be rejected based on summary statistics or the histogram?**

Answers may vary, but if the data shows a significant number of responses above 90 minutes, the null hypothesis could be rejected.

---

### Question 7

**State the null and alternative hypotheses:**

$$H_0: \mu = 90$$  
$$H_A: \mu > 90$$  

Where $\mu$ is the mean amount of TV watched per day by OSU students.

---

### Question 8

**Report the t-statistic and degrees of freedom:**

```{r}

t.test(tvDs$minutes, conf.level = 0.90, mu = 90, alternative = "greater")

result <- t.test(tvDs$minutes, conf.level = 0.90, mu = 90)



conf.
# Hypothesis test parameters
mu_0 <- 90  # Null hypothesis mean

# Calculate the t-statistic
t_stat <- (mean_value - mu_0) / (standard_deviation / sqrt(sample_size))

# Degrees of freedom
df <- sample_size - 1

# Print the t-statistic and degrees of freedom
print(paste("t-statistic:", round(t_stat, 5)))
print(paste("Degrees of freedom:", df))
```

**Results:**

$$t_{100} = -0.94003$$  
Degrees of freedom: 100


# Conclusion and Confidence Interval

### Question 9

```{r}
# Calculate the p-value for a one-sided test (alternative: mean > 90)
p_value <- pt(t_stat, df, lower.tail = FALSE)

# Print the p-value
print(paste("p-value:", round(p_value, 5)))
```

**Result:**

p-value: `0.82527`

There is no evidence to suggest that OSU students watch more than 90 minutes of TV per day based on the available data.

---

### Question 10

**Report the 90% confidence interval for $\mu$:**

```{r}
# Confidence level
conf_level <- 0.90
sd <- sd(tv_data$minutes, na.rm = TRUE)


# Margin of error for the confidence interval
t_cric <- qt((1 + conf_level) / 2, df)
ME <- t_cric * (sd / sqrt(n))

# Confidence interval bounds
ci_lower <- mean_value - ME
ci_upper <- mean_value + ME

# Print the confidence interval
print(paste("90% Confidence Interval: (", round(ci_lower, 2), "minutes, ", round(ci_upper, 2), "minutes)", sep = ""))
```

**Result:**

(53.49 minutes, 100.11 minutes)


### Question 11

**Interpret the confidence interval:**

We are 90% confident that the mean time OSU students spend watching TV per day is between 53.49 and  73.52 minutes.