---
title: "ST552 Homework 1"
author: 
    - "Brian Cervantes Alvarez"
date: "01-19-2023"
format: OSULatexTheme-pdf
execute: 
  warning: false
  message: false
editor: visual
---

# Problem 1:

## Part A: Writing reproducible code

```{r}
# Create a random dataset
set.seed(123) 
carDs <- data.frame(
  name = c("Toyota", "Honda", "Ford", "Chevrolet", "BMW"),
  mileage = c(30, 28, 25, 22, 20),
  color = c("Blue", "Red", "Silver", "Black", "White"),
  transmission = factor(c("Automatic", "Manual", "Automatic", "Automatic", "Manual"))
)
```

{{< pagebreak >}}

## Part B: Make A Plot
```{r}
library(ggplot2)
ggplot(carDs, aes(x = mileage, y = color)) +
  geom_point() +
  labs(title = "Mileage vs. Car Color",
       x = "Mileage",
       y = "Color")
```

{{< pagebreak >}}

## Part C: Load Mammals Data and Fit Linear Regression Model

```{r}
library(MASS)
data("mammals")
model <- lm(body ~ brain, data = mammals)
summary(model)
```

## Part D: Extract The Estimate of The Intercept

```{r}
interceptEstimate <- coef(model)[1]
print(paste0("The estimate of β0 is: ", interceptEstimate))
```

## Part E: Extract the estimate of sigma

```{r}
stdEstimate <- summary(model)$sigma
print(paste0("The estimate of σ is: ", stdEstimate))
```


{{< pagebreak >}}

# Problem 2: Simple Linear Regression

## Part A: Explanatory Variable and Response

The explanatory variable is Body weight (Bwt), and the response variable is Heart weight (Hwt).

## Part B: Scatterplot and Description

```{r}
data("cats")
plot(cats$Bwt, cats$Hwt, main = "Scatterplot of Body Weight vs. Heart Weight",
     xlab = "Body Weight (kg)", ylab = "Heart Weight (g)")
```

## Part C: Appropriateness of Linear Regression

The plot demonstrates that a linear regression model could be appropriate as there appears to be a linear trend between body weight and heart weight.

{{< pagebreak >}}

## Part D: Fit Linear Regression Model and Add Regression Line

```{r}
# Fit linear regression model
catModel <- lm(Hwt ~ Bwt, data = cats)

summary(catModel)

# Extract parameter estimates
intEstimate <- coef(catModel)[1]
slopeEstimate <- coef(catModel)[2]

print(paste0("Intercept estimate: ", round(intEstimate, 4)))
print(paste0("Slope estimate: ", round(slopeEstimate, 4)))

# Add regression line to the scatterplot
plot(cats$Bwt, cats$Hwt, main = "Scatterplot of Body Weight vs. Heart Weight",
     xlab = "Body Weight (kg)", ylab = "Heart Weight (g)")
abline(catModel, col = "red")

# Create Q-Q plot for the residuals
qqnorm(residuals(catModel), main = "Q-Q Plot for Residuals")
qqline(residuals(catModel), col = "red", lty = 2) # Reference Line

```

{{< pagebreak >}}

## Part E: Conclusion About The Relationship

The linear regression analysis indicates a statistically significant positive relationship between cat body weight and heart weight, with a slope estimate of $4.0341$. Approximately $64.41\%$ of the variability in heart weight can be explained by the linear relationship with body weight.

## Part F: Hypothesis Test for Slope Coefficient

The highly significant slope estimate of $4.0341$ in the linear regression model with the $pval < 2.2e-16$ indicates a substantial and practically meaningful positive relationship between cat body weight and heart weight.

```{r}
# Test hypothesis
test <- summary(catModel)$coefficients[2, c("Estimate",
                                            "Std. Error",
                                            "t value",
                                            "Pr(>|t|)")]
results <- data.frame(
  Estimate = test["Estimate"],
  `Std. Error` = test["Std. Error"],
  `t value` = test["t value"],
  `Pr(>|t|)` = test["Pr(>|t|)"]
)
results
```

## Part G: Predicted Mean Heart Weight for a Cat with Body Weight 3.5 Kilograms

```{r}
# Predict mean heart weight
predHeartWeight <- predict(catModel, 
                           data.frame(Bwt = 3.5), 
                           interval = "confidence")
print(paste0("Predicted mean heart weight for a cat with body weight 3.5kg: ",
    round(predHeartWeight[1],2), 
    "g"))
```

